### 2024-09-21 cognitive-bias.qmd
20240921 - base rates

Prospect Theory and Loss Aversion: People Tend to Weigh Losses More Heavily Than Gains. Adapted from <https://commons.wikimedia.org/wiki/File:Graphique_th%C3%A9orie_des_perspectives.svg>.
:::

@@

The *base rate fallacy* (aka base rate neglect) is the tendency to ignore information about the general probability of an event in favor of specific information about the event.
The *base rate* is a marginal probability, which is the general probability of an event irrespective of other things.
Focusing on the specific information about the event (and thus underweighting the base rate) involves the [representativeness heuristic](#sec-heuristicsRepresentativeness) [@Kahneman2011]—making judgments based on how similar the event is to a mental prototype regardless of how frequent the event the event actually is.
As noted by @Kahneman2011, "people who have information about an individual case rarely feel the need to know the statistics of the class which the case belongs." (p. 249).
In general, people tend to overestimate the likelihood of low base-rate events [@Kahneman2011].

@@

>>>>

Prospect Theory and Loss Aversion: People Tend to Weigh Losses More Heavily Than Gains. That is, the utility function is steeper in the loss (red) region than in the gain (blue) region. Adapted from <https://commons.wikimedia.org/wiki/File:Graphique_th%C3%A9orie_des_perspectives.svg>.
:::

@@

The *base rate fallacy* (aka base rate neglect) is the tendency to ignore information about the general probability of an event in favor of specific information about the event.
The *base rate* is a marginal probability, which is the general probability of an event irrespective of other things.
The, a "low base-rate" event is an event with a low probability of occurring (i.e., an unlikely event).
Focusing on the specific information about the event (and thus underweighting the base rate) involves the [representativeness heuristic](#sec-heuristicsRepresentativeness) [@Kahneman2011]—making judgments based on how similar the event is to a mental prototype regardless of how frequent the event the event actually is.
As noted by @Kahneman2011, "people who have information about an individual case rarely feel the need to know the statistics of the class which the case belongs." (p. 249).
In general, people tend to a) overestimate the likelihood of low base-rate events and b) overweight low base-rate events in their decisions [@Kahneman2011].
This is especially true if the unlikely event attracts attention, because of the [confirmatory bias](#sec-cognitiveBiasesConfirmation) of memory [@Kahneman2011].

@@

The [framing](#sec-cognitiveBiasesFraming) of an event can also alter its weighting.
For instance, concrete representations tend to be weighted more heavily than abstract representations [@Kahneman2011].
Abstract representations are framed in terms of "how likely" the event will occur in in terms of chance, risk, probability, or likelihood.
Concrete representations are framed in terms of  "how many" people will be affected by the event.
An example abstract representation is that there is a 0.001% chance of dying from this particular disease.
When framed in this way, the risk appears small.
Now, consider the same information framed in terms of a concrete representation: 1 in 100,000 people will die from this disease.
Due to a phenomenon called *denominator neglect*, low base-rate events are more heavily weighted when they are framed using a concrete representation than an abstract representation [@Kahneman2011].

There are low base-rate events that tend to be underweighted when making decisions, such as whether to make anticipatory preparations for a future earthquake, housing bubble, or financial crisis [@Kahneman2011].
One major cause of underweighting is if a person has never experienced the rare event (i.e., the earthquake or finanical crisis) [@Kahneman2011].
However, for low base-rate events that people have experienced, the [availability](#sec-heuristicsAvailability) and [representativeness](#sec-heuristicsRepresentativeness) heuristics tend to lead people to overestimate the likelihood that the unlikely event occurs.
In general, many factors—including obsessive concerns about the event, vivid images of the event, concrete representations of the consequences of the event (e.g., "a disease kills 1 person out of every 1,000"), and explicit reminders of the event—contribute to the overweighting of unlikely events [@Kahneman2011].
